---
title: "Reading and writing HDF5 files with R"
author: "Travis Hinkelman"
date: "2020-08-13"
output: ioslides_presentation
widescreen: true
css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## [HDF5](https://www.hdfgroup.org/solutions/hdf5/)

- High performance data software library and file format
- Manage, process, and store heterogeneous data
- Built for fast I/O processing and storage

## [DSM2](https://water.ca.gov/Library/Modeling-and-Analysis/Bay-Delta-Region-models-and-tools/Delta-Simulation-Model-II)

- 1D mathematical model for dynamic simulation of 1D hydrodynamics, water quality, and particle tracking in a network of riverine or estuarine channels.

- Widely used for planning and management of the Delta

- HDF5 is one of file formats used for DSM2 output

## [rhdf5](https://www.bioconductor.org/packages/release/bioc/html/rhdf5.html)

- Available through Bioconductor

- [Rhdf5lib](https://www.bioconductor.org/packages/release/bioc/html/Rhdf5lib.html) provides C and C++ HDF5 libraries

```{r echo = TRUE, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("rhdf5")
```

## File Contents

```{r echo = TRUE}
library(rhdf5)

path = "sjr1500_omr5000.h5"
head(h5ls(path), 5)
```

```{r echo = TRUE}
unique(h5ls(path)$group)
```

## Read Attributes

```{r echo = TRUE}
h5readAttributes(path, "hydro/data/channel flow")
```

## Generate Dates

```{r echo = TRUE}
hydro_attr <- h5readAttributes(path, "hydro")
flow_attr <- h5readAttributes(path, "hydro/data/channel flow")

num_int <- hydro_attr$`Number of intervals`
time_int <- hydro_attr$`Time interval`
start_dt <- flow_attr$start_time

datetimes <- seq.POSIXt(from = as.POSIXct(start_dt, tz = "UTC"),
                        by = paste(time_int, "mins"),
                        length.out = num_int)

head(datetimes)
```

## Read Data

```{r echo = TRUE}
h5read(path, "hydro/geometry/channel_number")[90:115]

h5read(path, "hydro/geometry/channel_location")

trimws(h5read(path, "hydro/geometry/channel_location"))
```

## Calculate Velocity

```{r echo = TRUE}
flow <- h5read(path, "hydro/data/channel flow")
dim(flow)

area <- h5read(path, "hydro/data/channel area")
dim(area)

velocity <- flow/area
```

## Array Dimensions

- 1st dimension is channel location (1 = upstream, 2 = downstream)
- 2nd dimension is channel number
- 3rd dimension is interval number

```{r echo = TRUE}
velocity[1, 1:5, 1:5]
```

If you are using HDFView, indexes start at 0, not 1.