---
title: "Reading and writing HDF5 files with R"
author: "Travis Hinkelman"
date: "8/13/2020"
output: ioslides_presentation
css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## [HDF5](https://www.hdfgroup.org/solutions/hdf5/)

- High performance data software library and file format
- Manage, process, and store heterogeneous data
- Built for fast I/O processing and storage

## [DSM2](https://water.ca.gov/Library/Modeling-and-Analysis/Bay-Delta-Region-models-and-tools/Delta-Simulation-Model-II)

- 1D mathematical model for dynamic simulation of 1D hydrodynamics, water quality, and particle tracking in a network of riverine or estuarine channels.

- Widely used for planning and management of the Delta

- HDF5 is one of file formats used for DSM2 output

## [rhdf5](https://www.bioconductor.org/packages/release/bioc/html/rhdf5.html)

- Available through Bioconductor

- Takes a dependency on [Rhdf5lib](https://www.bioconductor.org/packages/release/bioc/html/Rhdf5lib.html), which provides C and C++ HDF5 libraries

```{r echo = TRUE, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("rhdf5")
```

## List File Contents
```{r echo = TRUE}
library(rhdf5)
path = "sjr1500_omr5000.h5"
fc <- h5ls(path)
```
