---
title: "Reading and writing HDF5 files with R"
author: "Travis Hinkelman"
date: "2020-08-13"
output: ioslides_presentation
widescreen: true
css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## [HDF5](https://www.hdfgroup.org/solutions/hdf5/)

- High performance data software library and file format
- Manage, process, and store heterogeneous data
- Built for fast I/O processing and storage

## [DSM2](https://water.ca.gov/Library/Modeling-and-Analysis/Bay-Delta-Region-models-and-tools/Delta-Simulation-Model-II)

- 1D mathematical model for dynamic simulation of 1D hydrodynamics, water quality, and particle tracking in a network of riverine or estuarine channels.

- Widely used for planning and management of the Delta

- HDF5 is one of file formats used for DSM2 output

## [rhdf5](https://www.bioconductor.org/packages/release/bioc/html/rhdf5.html)

- Available through Bioconductor

- [Rhdf5lib](https://www.bioconductor.org/packages/release/bioc/html/Rhdf5lib.html) provides C and C++ HDF5 libraries

```{r echo = TRUE, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("rhdf5")
```

## File Contents

```{r echo = TRUE}
library(rhdf5)

path = "sjr1500_omr5000.h5"
head(h5ls(path), 5)
```

```{r echo = TRUE}
unique(h5ls(path)$group)
```

## Read Attributes

```{r echo = TRUE}
h5readAttributes(path, "hydro/data/channel flow")
```

## Generate Dates

```{r echo = TRUE}
hydro_attr <- h5readAttributes(path, "hydro")
flow_attr <- h5readAttributes(path, "hydro/data/channel flow")

num_int <- hydro_attr$`Number of intervals`
time_int <- hydro_attr$`Time interval`
start_dt <- flow_attr$start_time

datetimes <- seq.POSIXt(from = as.POSIXct(start_dt, tz = "UTC"),
                        by = paste(time_int, "mins"),
                        length.out = num_int)

head(datetimes)
```

## Read Data

```{r echo = TRUE}
h5read(path, "hydro/geometry/channel_number")[90:115]

h5read(path, "hydro/geometry/channel_location")

trimws(h5read(path, "hydro/geometry/channel_location"))
```

## Read Data

```{r echo = TRUE}
flow_attr <- h5read(path, "hydro/data/channel flow", read.attributes = TRUE)
str(flow_attr)
attr(flow_attr, "start_time")
```

## Calculate Velocity

```{r echo = TRUE}
flow <- h5read(path, "hydro/data/channel flow")
dim(flow)

area <- h5read(path, "hydro/data/channel area")
dim(area)

velocity <- flow/area
```

## Array Dimensions

- 1st dimension is channel location (1 = upstream, 2 = downstream)
- 2nd dimension is channel number
- 3rd dimension is interval number

```{r echo = TRUE}
velocity[1, 1:5, 1:5]
```

If you are using HDFView, indexes start at 0, not 1.

## Read Performance

```{r echo = TRUE, eval = FALSE}
pre_slice <- function(dim1, dim2, dim3){
  h5read(path, "hydro/data/channel flow", index = list(dim1, dim2, dim3))
  h5closeAll() }

post_slice <- function(dim1, dim2, dim3){
  h5read(path, "hydro/data/channel flow")[dim1, dim2, dim3, drop = FALSE]
  h5closeAll() }

microbenchmark::microbenchmark(
  pre_slice(1, 1:52, 1:144), 
  post_slice(1, 1:52, 1:144),
  pre_slice( 1:2, 1:465, 1:1297),
  post_slice(1:2, 1:465, 1:1297),
  pre_slice( 1, sample(1:517, 52), sample(1:1441, 144)), 
  post_slice(1, sample(1:517, 52), sample(1:1441, 144)),
  pre_slice( 1, sample(1:517, 104), sample(1:1441, 288)), 
  post_slice(1, sample(1:517, 104), sample(1:1441, 288)),
  times = 25L)
```

## Read Performance

```
Unit: milliseconds
                                                   expr       min    median       max
                              pre_slice(1, 1:52, 1:144)  15.44713  17.48160  33.19277
                             post_slice(1, 1:52, 1:144)  82.48233  88.14472 137.34435
                          pre_slice(1:2, 1:465, 1:1297)  77.68968  89.61094 239.75474
                         post_slice(1:2, 1:465, 1:1297)  98.44346 110.17357 136.78327
   pre_slice(1, sample(1:517, 52), sample(1:1441, 144)) 120.83129 134.07072 182.41477
  post_slice(1, sample(1:517, 52), sample(1:1441, 144))  82.65552  91.79320 152.53498
  pre_slice(1, sample(1:517, 104), sample(1:1441, 288)) 375.06928 424.33413 740.61241
 post_slice(1, sample(1:517, 104), sample(1:1441, 288))  78.35426  89.60911 107.31904
```